```{r load_packages, message= FALSE, warning = FALSE}
library(GEOquery)
library(R.utils)
library(gridExtra)
library(reshape2)
library(ggplot2)
library(limma)
library(dplyr)
library(DT)
library(plotly)
library(stringr)
library(tidyr)
library(DT)
library(reshape2)
library(RColorBrewer)
library(cvTools)
library(stringr)
library(FSelector)
library(ada)
library(caret)
library(pROC)
```

```{r gse1}

gse1 <- getGEO('GSE28146')
gse1 <- gse1$GSE28146_series_matrix.txt.gz


```

```{r features1}
  
#### Features ####

features1 <- gse1@featureData@data

#isolating gene_codes

features1 <- features1 %>% mutate(gene_codes = str_trim(str_extract(.[["Gene Symbol"]], "^[^/]+")))

#summarising the featdf to only the important variables 

features1 <- features1 %>% select(GB_ACC,ID,gene_codes) #test models with the remaining 3 columns left out

############################
```

```{r Phenotypes1}


#### Phenotypes ####


phenotypes1 <- gse1@phenoData@data

#sex 

phenotypes1$sex <- str_extract(phenotypes1$characteristics_ch1, "(?<=Sex: ).*")

#age cleaning 

phenotypes1$age <- as.numeric(str_extract(phenotypes1$characteristics_ch1.1, "[0-9]+"))

#outcome

phenotypes1$outcome <- str_extract(phenotypes1$characteristics_ch1.2, "(?<=disease status: ).*")

phenotypes1$outcome <- ifelse(
  str_detect(phenotypes1$characteristics_ch1.2, "Control"),
  "Control",
  "AD"
)

#Simplify the frame
phenotypes1 <- phenotypes1 %>% select(geo_accession,age,sex,outcome)

```

```{r eMat1}
#eMat

eMat1 <- exprs(gse1)
rownames(eMat1) <- features1$gene_codes

```


```{r LoadingGSE2}
gse2 <- getGEO("GSE66351")
gse2 <- gse2$GSE66351_series_matrix.txt.gz
```


```{r features2}
  
#### Features ####

features2 <- gse2@featureData@data

#gene_codes and cleaning
features2$gene_codes <- gsub(";.*", "", features2$UCSC_RefGene_Name)

features2 <- features2[!is.na(features2$gene_codes), ]
features2 <- features2[!is.na(features2$gene_codes) & features2$gene_codes != "", ]

#selecct important
features2 <- features2 %>% select(ID,gene_codes) 

############################
```

```{r Phenotypes2}


#### Phenotypes ####

  
phenotypes2 <- gse2@phenoData@data

#sex 

phenotypes2$sex <- str_extract(phenotypes2$characteristics_ch1.5, "(?<=Sex: ).*")


phenotypes2$sex <- ifelse(
  str_detect(phenotypes2$sex, "F"),
  "Female",
  "Male"
)


#age cleaning 

phenotypes2$age <- as.numeric(phenotypes2$`age:ch1`)

#outcome


phenotypes2$outcome <- ifelse(
  str_detect(phenotypes2$'diagnosis:ch1', "CTRL"),
  "Control",
  "AD"
  
)


#Control      AD 
#     92     128 


#Simplify the frame
phenotypes2 <- phenotypes2 %>% select(geo_accession,age,sex,outcome)

```

```{r eMat2}
#eMat

eMat2 <- exprs(gse2)
eMat2 <- eMat2[features2$ID, ]
rownames(eMat2) <- features2$gene_codes

```


```{r COMBINE}

combined_features <- bind_rows(features1, features2)

combined_phenotypes <- bind_rows(phenotypes1, phenotypes2)

common_genes <- intersect(rownames(eMat1), rownames(eMat2))
eMat1_common <- eMat1[common_genes, ]
eMat2_common <- eMat2[common_genes, ]
combined_eMat <- cbind(eMat1_common, eMat2_common)

```

```{r outcomeasfactor}
# First, check and fix the outcome variable
print("Structure of combined phenotypes:")
str(combined_phenotypes)

# Convert outcome to factor with proper levels
if (!is.factor(combined_phenotypes$outcome)) {
  print("Converting outcome to factor with Control as reference level")
  combined_phenotypes$outcome <- factor(combined_phenotypes$outcome, levels=c("Control", "AD"))
}

# Verify the factor levels
print("Levels of outcome variable after conversion:")
print(levels(combined_phenotypes$outcome))

# Check class balance
print("Class distribution:")
table(combined_phenotypes$outcome)
```

Limma method: Using differential expression analysis with limma to find genes that differ significantly between AD and Control groups.

Information Gain method: Using information theory to identify genes with the highest information content for distinguishing between outcome classes.

PCA method: Using Principal Component Analysis to find genes that contribute most to the top principal components (the most variance in the data).

```{r Top100GenesLimma}

# Using limma for differential expression

# Make sure outcome is a factor with Control as reference level
outcome <- factor(combined_phenotypes$outcome, levels=c("Control", "AD"))

# Create design matrix
design <- model.matrix(~outcome)

# Fit model
fit <- lmFit(combined_eMat, design)
fit <- eBayes(fit)

# Get top differentially expressed genes
results <- topTable(fit, coef="outcomeAD", number=100, sort.by="p")
top_limma_genes <- rownames(results)

# Create matrix with selected genes, transpose so samples are rows
X <- t(combined_eMat[top_limma_genes, ])
X_scaled <- scale(X)  # Scale the data *STANDARDISES FOR u=0, stdev = 1

head(top_limma_genes)
results
```

```{r Top100GenesInfoGain}

library(FSelector)

# Convert expression data to data frame with outcome
df <- data.frame(t(combined_eMat)) #IF YOU HAVENT ALREADY TRANSPOSED!!!
df$outcome <- outcome #MAKE SURE THIS IS SET AS FACTOR 

# Calculate information gain
weights <- information.gain(outcome ~ ., df)
top_ig_genes <- rownames(combined_eMat)[head(order(weights$attr_importance, decreasing=TRUE), 100)]

```

```{r Top100GenesPCA}
print("Performing PCA analysis...")
# PCA
pca_result <- prcomp(t(combined_eMat), scale=TRUE)

# Extract top genes contributing to first few PCs
loadings <- abs(pca_result$rotation[,1:3])
contrib <- rowSums(loadings)
top_pca_genes <- names(sort(contrib, decreasing=TRUE))[1:100]

print("Top 10 PCA genes:")
print(head(top_pca_genes, 10))
print("Contribution values of top 10 PCA genes:")
print(head(sort(contrib, decreasing=TRUE), 10))
```

```{r limmaADABOOST}
library(adabag)
library(caret)

# Verify combined phenotypes structure
print(dim(combined_phenotypes))  # Expect 220 rows, check columns
print(str(combined_phenotypes$outcome))  # Ensure "Control" and "AD" are recognized correctly
print(table(combined_phenotypes$outcome))  # Check class distribution

# Prepare data
X <- t(combined_eMat[top_limma_genes, ])  # Transpose so samples are rows
X_scaled <- scale(X)  # Standardize for better model performance
y <- combined_phenotypes$outcome  # Outcome labels

# Convert outcome to numeric (required for AdaBoost)
y_numeric <- as.numeric(y) - 1  # Convert "Control"=0, "AD"=1
print(length(y_numeric))  # Ensure correct number of samples

# Split data (80% train, 20% test)
set.seed(42)
sample_size <- floor(0.8 * length(y_numeric))
train_indices <- sample(seq_len(length(y_numeric)), size=sample_size)

X_train <- X_scaled[train_indices, ]
y_train <- y_numeric[train_indices]
X_test <- X_scaled[-train_indices, ]
y_test <- y_numeric[-train_indices]

# Verify split integrity
print(dim(X_train))  # Ensure samples and features match
print(length(y_train))  # Should match X_train rows
print(table(y_train))  # Check class balance
print(table(y_test))  # Check test set distribution

# Prepare training data
train_data <- data.frame(label = factor(y_train, levels = c(0,1)), X_train)

# Train AdaBoost model
boost_model <- boosting(label ~ ., data=train_data, mfinal=50)

# Predictions
pred <- predict(boost_model, newdata=data.frame(X_test))

# Evaluate model performance
conf_matrix <- confusionMatrix(factor(pred$class), factor(y_test))
print(conf_matrix)
```


```{r }
library(adabag)
library(caret)

# Verify combined phenotypes structure
print(dim(combined_phenotypes))  # Expect 220 rows, check columns
print(str(combined_phenotypes$outcome))  # Ensure "Control" and "AD" are recognized correctly
print(table(combined_phenotypes$outcome))  # Check class distribution

# Prepare data
X <- t(combined_eMat[top_ig_genes, ])  # Transpose so samples are rows
X_scaled <- scale(X)  # Standardize for better model performance
y <- combined_phenotypes$outcome  # Outcome labels

# Convert outcome to numeric (required for AdaBoost)
y_numeric <- as.numeric(y) - 1  # Convert "Control"=0, "AD"=1
print(length(y_numeric))  # Ensure correct number of samples

# Split data (80% train, 20% test)
set.seed(42)
sample_size <- floor(0.8 * length(y_numeric))
train_indices <- sample(seq_len(length(y_numeric)), size=sample_size)

X_train <- X_scaled[train_indices, ]
y_train <- y_numeric[train_indices]
X_test <- X_scaled[-train_indices, ]
y_test <- y_numeric[-train_indices]

# Verify split integrity
print(dim(X_train))  # Ensure samples and features match
print(length(y_train))  # Should match X_train rows
print(table(y_train))  # Check class balance
print(table(y_test))  # Check test set distribution

# Prepare training data
train_data <- data.frame(label = factor(y_train, levels = c(0,1)), X_train)

# Train AdaBoost model
boost_model <- boosting(label ~ ., data=train_data, mfinal=50)

# Predictions
pred <- predict(boost_model, newdata=data.frame(X_test))

# Evaluate model performance
conf_matrix <- confusionMatrix(factor(pred$class), factor(y_test))
print(conf_matrix)
```

```{r pcaADABOOST}
library(adabag)
library(caret)

# Verify combined phenotypes structure
print(dim(combined_phenotypes))  # Expect 220 rows, check columns
print(str(combined_phenotypes$outcome))  # Ensure "Control" and "AD" are recognized correctly
print(table(combined_phenotypes$outcome))  # Check class distribution

# Prepare data
X <- t(combined_eMat[top_pca_genes, ])  # Transpose so samples are rows
X_scaled <- scale(X)  # Standardize for better model performance
y <- combined_phenotypes$outcome  # Outcome labels

# Convert outcome to numeric (required for AdaBoost)
y_numeric <- as.numeric(y) - 1  # Convert "Control"=0, "AD"=1
print(length(y_numeric))  # Ensure correct number of samples

# Split data (80% train, 20% test)
set.seed(42)
sample_size <- floor(0.8 * length(y_numeric))
train_indices <- sample(seq_len(length(y_numeric)), size=sample_size)

X_train <- X_scaled[train_indices, ]
y_train <- y_numeric[train_indices]
X_test <- X_scaled[-train_indices, ]
y_test <- y_numeric[-train_indices]

# Verify split integrity
print(dim(X_train))  # Ensure samples and features match
print(length(y_train))  # Should match X_train rows
print(table(y_train))  # Check class balance
print(table(y_test))  # Check test set distribution

# Prepare training data
train_data <- data.frame(label = factor(y_train, levels = c(0,1)), X_train)

# Train AdaBoost model
boost_model <- boosting(label ~ ., data=train_data, mfinal=50)

# Predictions
pred <- predict(boost_model, newdata=data.frame(X_test))

# Evaluate model performance
conf_matrix <- confusionMatrix(factor(pred$class), factor(y_test))
print(conf_matrix)
```
